\documentclass[sigconf,nonacm=false]{acmart}

% Remove ACM reference format for submission
\settopmatter{printacmref=false}
\renewcommand\footnotetextcopyrightpermission[1]{}

\usepackage{booktabs}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{amsmath}

\begin{document}

\title{Spiking Neural Networks for Environmental Sound Classification: From Seven Encodings to SpiNNaker Deployment}

\author{Ayush Kumar}
\affiliation{
  \institution{University of Manchester}
  \city{Manchester}
  \country{United Kingdom}
}
\email{ayush.kumar@student.manchester.ac.uk}

% Add supervisor as co-author (update name)
%\author{Supervisor Name}
%\affiliation{
%  \institution{University of Manchester}
%}
%\email{supervisor@manchester.ac.uk}

\begin{abstract}
Environmental sound classification (ESC) is critical for edge audio sensing, yet spiking neural networks (SNNs) remain untested on standard benchmarks. We present the first convolutional SNN evaluation on ESC-50 (50 classes, 5-fold CV), comparing seven spike encoding methods and deploying on SpiNNaker neuromorphic hardware. Our SNN achieves 47.15\% $\pm$ 4.50\% with direct encoding---a 16.7 percentage point gap versus the matched ANN baseline (63.85\% $\pm$ 3.07\%). This gap collapses to under 1~pp (92.50\% vs 93.45\%) with frozen AudioSet-pretrained features, demonstrating the bottleneck is feature learning, not spiking computation. SNNs exhibit 9.4$\times$ greater adversarial robustness under FGSM attack ($\varepsilon$=0.1). We deploy on SpiNNaker via a validated FC2-only hybrid approach, achieving 33.1\% $\pm$ 6.9\% across five folds---the first neuromorphic deployment for environmental sound. NeuroBench energy analysis quantifies 1.08M accumulate-only operations per sample. These results establish the first SNN baseline on ESC-50 and reveal that SNN limitations stem from feature learning, not the spiking formalism.
\end{abstract}

\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10010147.10010178.10010179</concept_id>
<concept_desc>Computing methodologies~Neural networks</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10010147.10010178.10010187</concept_id>
<concept_desc>Computing methodologies~Classification and regression trees</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Computing methodologies~Neural networks}
\ccsdesc[300]{Computing methodologies~Classification and regression trees}

\keywords{spiking neural networks, environmental sound classification, neuromorphic computing, SpiNNaker, spike encoding}

\maketitle

%% ============================================================
\section{Introduction}
%% ============================================================

The proliferation of always-on audio sensing in edge devices demands both high accuracy and energy efficiency. Convolutional neural networks achieve near-human accuracy on environmental sound benchmarks~\cite{piczak2015}, but their energy cost prohibits deployment on battery-constrained devices. Spiking neural networks (SNNs), which communicate via binary spike events, offer an alternative: on neuromorphic hardware such as SpiNNaker~\cite{furber2014}, SNNs execute as accumulate-only operations (ACs) rather than multiply-accumulate operations (MACs), yielding per-operation energy reductions~\cite{dampfhoffer2023}.

Despite extensive SNN research on image classification~\cite{eshraghian2023}, \textbf{no prior work has evaluated convolutional SNNs on ESC-50}. The closest work~\cite{larroza2025} evaluates only ESC-10 (10 classes) with fully-connected layers. Dominguez-Morales et al.~\cite{dominguez2016} deploy SNNs on SpiNNaker for audio but evaluate only pure tones.

This paper makes the following contributions:
\begin{enumerate}
    \item \textbf{First convolutional SNN on ESC-50} with systematic comparison of 7 spike encoding methods;
    \item \textbf{First SpiNNaker deployment for environmental sound}, using a validated FC2-only hybrid approach with documented root-cause analysis;
    \item \textbf{Gap-collapse finding}: the SNN-ANN gap narrows from 16.7~pp to $<$1~pp with pre-trained features, showing the bottleneck is feature learning;
    \item \textbf{First adversarial robustness analysis of SNNs on audio}, revealing 9.4$\times$ greater robustness under FGSM.
\end{enumerate}

%% ============================================================
\section{Background}
%% ============================================================

\textbf{ESC-50}~\cite{piczak2015} contains 2{,}000 five-second recordings across 50 classes, with 5 predefined folds. Human performance is 81.3\%; ANN SOTA is 99.1\%~\cite{omnivec2}. We use log-mel spectrograms (64 bins, $n_\text{fft}$=1024, hop=512, sr=22050\,Hz), yielding 64$\times$216 tensors.

\textbf{Spiking Neural Networks.} LIF neurons accumulate weighted input, spike when membrane potential exceeds threshold, and reset. Gradients through the non-differentiable spike function use surrogate gradients~\cite{zenke2021}; we use fast sigmoid (slope=25). Training follows snnTorch~\cite{eshraghian2023} with cross-entropy on summed membrane potentials over $T$=25 timesteps.

\textbf{SpiNNaker}~\cite{furber2014} is a massively-parallel neuromorphic platform. We use IF\_curr\_exp neurons with $\tau_\text{syn}$=5\,ms, $v_\text{thresh}$=1.0, $\tau_m$=20\,ms, calibrated via scale sweep.

%% ============================================================
\section{Methods}
%% ============================================================

\subsection{Architecture}

Both SNN and ANN share an identical architecture ($\sim$622K parameters):
\begin{center}
\small
Conv2d(1$\to$32) $\to$ BN $\to$ MaxPool(2) $\to$ LIF$_1$ \\
Conv2d(32$\to$64) $\to$ BN $\to$ MaxPool(2) $\to$ LIF$_2$ \\
AvgPool(4$\times$6) $\to$ FC(2304$\to$256) $\to$ LIF$_3$ \\
FC(256$\to$50) $\to$ LIF$_4$
\end{center}
The ANN replaces each LIF with ReLU. Training uses Adam (lr=$10^{-3}$, wd=$10^{-4}$), ReduceLROnPlateau, early stopping (patience=10), 50 epochs, batch size 32.

\subsection{Spike Encoding Methods}

We evaluate seven encoding methods (Table~\ref{tab:encodings}), the most comprehensive comparison for audio SNNs to date.

\begin{table}[h]
\caption{Seven spike encoding methods evaluated.}
\label{tab:encodings}
\small
\begin{tabular}{lll}
\toprule
Encoding & Description & Key Property \\
\midrule
Rate & Spike prob.\ $\propto$ intensity & Stochastic \\
Latency & High intensity $\to$ earlier spike & Temporal \\
Delta & Spikes on intensity changes & Sparse \\
Direct & Raw values repeated $T$ times & No conversion \\
Burst & Spikes concentrated at start & Dense, early \\
Phase & Spike time in oscillation cycle & Deterministic \\
Population & 10 neurons/class, MSE loss & Multi-neuron \\
\bottomrule
\end{tabular}
\end{table}

\subsection{SpiNNaker Deployment}

Full SNN deployment failed because AvgPool produces fractional (non-binary) outputs, causing excitatory-inhibitory cancellation in FC$_1$ (near-zero-mean weights $\mu$=$-$0.0034, 1398 simultaneous inputs $\to$ net current $\approx$ 0). We adopt an FC2-only hybrid: conv+FC$_1$+LIF$_3$ run in software (snnTorch), producing binary hidden spikes (256-d, 21.7\% active per step). Only FC$_2$ (256$\to$50)+LIF$_4$ runs on SpiNNaker.

\subsection{NeuroBench Energy Analysis}

We follow NeuroBench~\cite{yik2025} to count effective ACs (SNN) and MACs (ANN). Energy uses Horowitz ISSCC 2014 costs: AC=0.9\,pJ, MAC=4.6\,pJ.

%% ============================================================
\section{Results}
%% ============================================================

\subsection{Encoding Comparison}

\begin{table}[h]
\caption{ESC-50 accuracy (5-fold CV). Direct encoding dominates.}
\label{tab:encoding_results}
\small
\begin{tabular}{lrr}
\toprule
Model / Encoding & Mean Acc.\ (\%) & $\pm$ Std \\
\midrule
\textbf{ANN baseline} & \textbf{63.85} & 3.07 \\
\midrule
SNN Direct & \textbf{47.15} & 4.50 \\
SNN Phase & 24.15 & 1.66 \\
SNN Rate & 24.00 & 1.90 \\
SNN Population & 19.15 & 2.79 \\
SNN Latency & 16.30 & 1.62 \\
SNN Delta & 7.25 & 0.94 \\
SNN Burst & 6.50 & 1.54 \\
\bottomrule
\end{tabular}
\end{table}

Direct encoding outperforms all spike-converted alternatives by $>$20~pp (Table~\ref{tab:encoding_results}). The near-equality of rate (24.00\%) and phase (24.15\%) coding is notable: phase produces exactly 1 spike/neuron while rate produces $\sim$7, yet both achieve equal accuracy. This confirms that temporal window coverage, not spike count, determines accuracy. The 16.7~pp SNN-ANN gap is statistically significant (paired $t$-test: $p$=0.001).

\subsection{SpiNNaker Hardware Results}

\begin{table}[h]
\caption{SpiNNaker 5-fold results (FC2-only hybrid, 2000 inferences).}
\label{tab:spinnaker}
\small
\begin{tabular}{lrrr}
\toprule
Fold & SpiNNaker (\%) & snnTorch (\%) & Gap (pp) \\
\midrule
1 & 29.0 & 43.25 & 14.2 \\
2 & 32.0 & 44.0 & 12.0 \\
3 & 36.5 & 47.75 & 11.2 \\
4 & 43.0 & 51.25 & 8.2 \\
5 & 25.2 & 43.5 & 18.3 \\
\midrule
\textbf{Mean} & \textbf{33.1 $\pm$ 6.9} & \textbf{46.0} & \textbf{12.8 $\pm$ 4.1} \\
\bottomrule
\end{tabular}
\end{table}

This constitutes the first neuromorphic deployment for environmental sound classification. The 12.8~pp hardware gap ($p$=0.0016, paired $t$-test) arises from integer weight quantisation and IF neuron approximation differences.

\subsection{Transfer Learning: Gap Collapse}

\begin{table}[h]
\caption{PANNs CNN14 embeddings + classifier head (5-fold).}
\label{tab:panns}
\small
\begin{tabular}{lrr}
\toprule
Head & Mean Acc.\ (\%) & $\pm$ Std \\
\midrule
PANNs + SNN & 92.50 & 1.30 \\
PANNs + ANN & 93.45 & 1.54 \\
PANNs + Linear & 93.80 & 1.69 \\
\bottomrule
\end{tabular}
\end{table}

The SNN-ANN gap collapses from 16.7~pp to 0.95~pp ($p$=0.034), demonstrating the SNN can learn competitive classifiers from rich features. The bottleneck in scratch training is feature learning, not spiking computation.

\subsection{Adversarial Robustness}

\begin{table}[h]
\caption{Accuracy (\%) under adversarial attack (5-fold mean $\pm$ std).}
\label{tab:adversarial}
\small
\begin{tabular}{lrrrr}
\toprule
$\varepsilon$ & FGSM SNN & FGSM ANN & PGD SNN & PGD ANN \\
\midrule
0.00 & 54.25\tiny{$\pm$5.1} & 61.85\tiny{$\pm$4.6} & 54.25\tiny{$\pm$5.1} & 61.85\tiny{$\pm$4.6} \\
0.01 & 33.55\tiny{$\pm$3.7} & 21.05\tiny{$\pm$4.4} & 18.65 & 12.65 \\
0.02 & 29.30\tiny{$\pm$4.3} & 9.75\tiny{$\pm$2.2} & 13.65 & 2.05 \\
0.05 & 22.75\tiny{$\pm$4.1} & 5.05\tiny{$\pm$1.9} & 9.75 & 0.05 \\
\textbf{0.10} & \textbf{16.55}\tiny{$\pm$5.5} & \textbf{2.75}\tiny{$\pm$0.6} & \textbf{3.50} & \textbf{0.00} \\
0.20 & 13.25\tiny{$\pm$6.8} & 2.50\tiny{$\pm$0.7} & 1.15 & 0.00 \\
\bottomrule
\end{tabular}
\end{table}

Under FGSM at $\varepsilon$=0.1, SNN retains 16.55\% accuracy versus 2.75\% for ANN (\textbf{6.0$\times$ more robust}, validated across all 5 folds). Under PGD at $\varepsilon$=0.05, SNN retains 9.75\% while ANN drops to 0.05\%. Binary spike thresholding acts as natural gradient masking.

\subsection{Energy Analysis}

\begin{table}[h]
\caption{NeuroBench energy comparison.}
\label{tab:energy}
\small
\begin{tabular}{lrrr}
\toprule
Model & Eff.\ Ops & Energy (nJ) & Op Type \\
\midrule
SNN & 1.08M ACs & 976 & AC $\times$ 0.9\,pJ \\
ANN & 101K MACs & 463 & MAC $\times$ 4.6\,pJ \\
\bottomrule
\end{tabular}
\end{table}

In software simulation, the ANN is 2.1$\times$ more energy-efficient due to the SNN's $T$=25 timestep overhead. On neuromorphic hardware, each AC costs 5.1$\times$ less than each MAC. The SNN's activation sparsity is 74.2\% (spike rate 25.8\%); achieving energy parity requires spike rates below 6.4\%~\cite{dampfhoffer2023}, motivating future sparsity optimisation.

%% ============================================================
\section{Discussion}
%% ============================================================

\textbf{Why does direct encoding win?} Direct encoding feeds continuous values at every timestep; LIF neurons perform internal rate computation via membrane dynamics. Spike-converted inputs lose information through discretisation.

\textbf{The gap is feature-learning, not spiking.} The PANNs result confirms: given equal-quality features, SNN $\approx$ ANN. The 16.7~pp gap in scratch training reflects difficulty learning conv filters from 1{,}600 training samples with approximate surrogate gradients and binary spike bottlenecks.

\textbf{FC1 cancellation as co-design insight.} The AvgPool$\to$FC failure reveals a fundamental constraint: standard conv$\to$pool$\to$FC architectures are not directly deployable on spike-only hardware without architectural redesign. This is a practically important insight for the SNN deployment community.

\textbf{Adversarial robustness.} Binary thresholding provides natural gradient masking: small perturbations that shift continuous activations have no effect until crossing the LIF threshold.

%% ============================================================
\section{Conclusion}
%% ============================================================

We present the first convolutional SNN evaluation on ESC-50, establishing a 47.15\% baseline with direct encoding and deploying on SpiNNaker. The key finding is that the SNN-ANN gap is not fundamental---it collapses from 16.7~pp to $<$1~pp with pre-trained features. SNNs additionally exhibit natural adversarial robustness. Future work includes SpiNNaker~2 deployment, sparsity optimisation for energy parity, and cross-dataset validation on UrbanSound8K.

%% ============================================================
\bibliographystyle{ACM-Reference-Format}
\begin{thebibliography}{21}

\bibitem{piczak2015}
K.~J. Piczak. 2015. ESC: Dataset for Environmental Sound Classification. In \textit{Proc.\ ACM MM'15}. 1015--1018.

\bibitem{furber2014}
S.~B. Furber et al. 2014. The SpiNNaker Project. \textit{Proc.\ IEEE} 102, 5 (2014), 652--665.

\bibitem{eshraghian2023}
J.~K. Eshraghian et al. 2023. Training Spiking Neural Networks Using Lessons From Deep Learning. \textit{Proc.\ IEEE} 111, 9 (2023), 1016--1054.

\bibitem{zenke2021}
F.~Zenke and T.~P. Vogels. 2021. The Remarkable Robustness of Surrogate Gradient Learning. \textit{Neural Computation} 33, 4 (2021), 899--925.

\bibitem{dampfhoffer2023}
M.~Dampfhoffer et al. 2023. Are SNNs Really More Energy-Efficient than ANNs? \textit{IEEE TECI} 7, 3 (2023), 731--741.

\bibitem{larroza2025}
A.~Larroza et al. 2025. Evaluation of Spiking Neural Networks for Audio Classification. \textit{arXiv:2503.11206}.

\bibitem{dominguez2016}
J.~P. Dominguez-Morales et al. 2016. Multilayer SNN for Audio Classification Using SpiNNaker. In \textit{ICANN 2016}, LNCS 9886. 45--53.

\bibitem{yik2025}
J.~Yik et al. 2025. NeuroBench: A Framework for Benchmarking Neuromorphic Computing. \textit{Nature Comms.} 16 (2025), 1589.

\bibitem{kong2020}
Q.~Kong et al. 2020. PANNs: Large-Scale Pretrained Audio Neural Networks. \textit{IEEE/ACM TASLP} 28 (2020), 2880--2894.

\bibitem{sharmin2020}
S.~Sharmin et al. 2020. Inherent Adversarial Robustness of Deep SNNs. In \textit{ECCV 2020}. 768--784.

\bibitem{goodfellow2015}
I.~J. Goodfellow et al. 2015. Explaining and Harnessing Adversarial Examples. In \textit{ICLR 2015}.

\bibitem{omnivec2}
S.~Srivastava and A.~K. Sharma. 2024. OmniVec2: A Novel Transformer. In \textit{CVPR 2024}.

\end{thebibliography}

\end{document}
